Towards Practical First-Order Model Counting

### 2024-04-12 doc/paper.tex
Add lots of references (and some other changes)

\newcommand{\Cranetwo}{\textsc{Crane}~2}

@@

% TODO: check if this follows the formatting guidelines
\author{%
Ananth K. Kidambi$^1$\and
Guramrit Singh$^1$\and
Paulius Dilkas$^2$\and
Kuldeep S. Meel$^3$ \\
\affiliations
$^1$Indian Institute of Technology Bombay, India\\
$^2$National University of Singapore, Singapore\\
$^3$University of Toronto, Canada\\
\emails
\{210051002, 210050061\}@iitb.ac.in,
paulius.dilkas@nus.edu.sg,
meel@cs.toronto.edu
}

@@

\paragraph{Papers to cite.}
\begin{itemize}
  \item relevant theoretical work~\cite{DBLP:conf/aaai/MalhotraS22} (there is
        more, e.g., the LICS 2018 paper)
  \item original domain recursion~\cite{DBLP:conf/nips/Broeck11}
  \item recent overview paper~\cite{DBLP:conf/ijcai/Kuzelka23}
  \item conjecturing recurrence relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}
  \item algorithms
  \begin{itemize}
    \item \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
    \item \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}
    \item \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}
    \item approximate~\cite{DBLP:conf/ijcai/BremenK20}
  \end{itemize}
  \item liftable fragments
  \begin{itemize}
    \item \Ctwo~\cite{DBLP:journals/jair/Kuzelka21}
    \item tree axioms~\cite{DBLP:journals/ai/BremenK23}
    \item linear order axioms~\cite{DBLP:conf/aaai/TothK23}
  \end{itemize}
  \item applications
  \begin{itemize}
    \item extensions to
          sampling~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}
    \item discovery of combinatorial
          sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23}
  \end{itemize}
\end{itemize}

\todo[inline]{Disambiguate between \protect\Crane, \textsc{Crane}, and
  \Cranetwo}

\paragraph{Changes to the compilation process.}
\begin{itemize}
  \item We make the FCGs satisfy the following property. For every domain
        $\Delta$, for every directed path $P$ without Ref nodes, the number of
        atom counting or (generalised) domain recursion or independent partial
        grounding operations on $\Delta$ on $P$ is at most one.
  \item when checking whether recursion is possible between a source and a
        target formula, the target formula (maybe the source formula as well?)
        must not include a domain as well as its subdomain. This might be
        equivalent to the formula being `unshattered'. Such formulas could be
        seen as transitory, and allowing such recursive calls leads to FCGs that
        I cannot make sense of.
\end{itemize}

@@

\paragraph{Three types of logics}
\begin{itemize}
  \item See \cref{tbl:logics} for a detailed comparison. The notation introduced
        in the table is standard for \Ctwo{}, new for \UFO{}, and redefined to
        be more specific for \FO{}.
  \item All three logics are function-free.
  \item Domains are always assumed to be finite.
  \item
        \begin{itemize}
          \item \FO{} is used as the input format for
                \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
                and its extensions
                \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}}~\cite{DBLP:conf/kr/DilkasB23}
                and \Cranetwo{}.
          \item \Ctwo{} is discussed in the literature on
                \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21} and related
                methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}
          \item \UFO{} is the input format supported by a version of
                \textsc{FastWFOMC} obtained directly from the authors. Note that
                the publicly available
                version\footnote{\url{https://www.comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
                does not support any cardinality constraints.
        \end{itemize}
  \item Note that, in the case of \textsc{ForcLift} and its extensions, support
        for a formula as valid input does not imply that the algorithm will be
        able to compile the formula into a circuit or graph suitable for lifted
        model counting. However, it is known that \textsc{ForcLift} compilation
        is guaranteed to succeed on any \FO{} formula without constants and with
        at most two variables~\cite{DBLP:conf/nips/Broeck11}.
\end{itemize}

@@

\begin{algorithm}[t]
  \caption{The core part of \Cranetwo{} that compiles a formula...}\label{alg:main}% TODO: start here
  \KwIn{formula $\phi$}
  \KwOut{set $\mathcal{E}$ of equations}
  $(\mathcal{E}, \mathcal{F}, \mathcal{D}) \gets \Crane{$\phi$}$\;
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$\mathcal{E}$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i}$ is a constant}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $(\mathcal{E}', \text{\textunderscore}, \text{\textunderscore}) \gets \Crane{$\psi$}$\;
    $\mathcal{E} \gets \mathcal{E} \cup \mathcal{E}'$\;
  }
\end{algorithm}

See \cref{alg:main}.

We write \Crane for a slight modification of the original \textsc{Crane} that
compiles a formula $\phi$ into:
\begin{itemize}
  \item a set $\mathcal{E}$ of equations,
  \item a map $\mathcal{F}$ from function names to formulas, and
  \item a map $\mathcal{D}$ from function names and parameter indices to
        domains.
\end{itemize}

% TODO: update: crane or crane2
The evaluation of base cases is done by simplifying the clauses and then using
\textsc{Crane} to find the base cases. A particular domain is selected, and the
clauses are simplified. Then, \textsc{Crane} is called on those clauses to
evaluate the base cases. After that, we change the function names and variable
to make it consistent with the previous domain to variable mapping, and append
these base cases to the set of equations.

\todo[inline,caption={}]{extend it to work with base cases that are themselves
  recursive}

@@

Comparing \textsc{Crane} and \textsc{FastWFOMC} on a larger set of benchmarks is
challenging because there is no automated way to translate a formula in \FO{} or
\Ctwo{} into \UFO{} (or even check if such an encoding is possible).

@@

\paragraph{Setup.}
\begin{itemize}
  \item The experiments were run on an AMD~Ryzen~7~5800H processor with
        \SI{16}{\gibi\byte} of memory and Arch Linux~6.8.2-arch2-1 operating
        system. \textsc{FastWFOMC} was run using Python~3.8.19 with
        Python-FLINT~0.5.0.
\end{itemize}

@@

\paragraph{Results.}
\begin{itemize}
  \item As shown in \cref{fig:plot}, the runtimes of all compilation-based
        algorithms remain practically constant in contrast to the rapidly
        increasing runtimes of \textsc{FastWFOMC}.
  \item Note that \textsc{Crane-BFS} is able to handle more instances than
        \textsc{ForcLift} (e.g., the permutation-counting problem in our
        experiments and other problems in my previous work).
  \item Although the search/compilation part is slower in \textsc{Crane} than in
        \textsc{ForcLift}, the difference is negligible.
  \item The runtimes of three out of four WFOMC algorithms appear constant
        because---for these counting problems and domain sizes---compilation
        time dominates inference time (recall that compilation time is
        independent of domain sizes). Indeed, the maximum inference time of both
        \textsc{Crane-BFS} and \textsc{Crane-Greedy} across these experiments is
        only \SI{4}{\milli\second}.
  \item The runtimes of \textsc{Crane} have lower variation than those of
        \textsc{ForcLift} because with \textsc{ForcLift} we compile the formula
        anew for each domain size whereas with \textsc{Crane} we compile it once
        and reuse the resulting C++ program for all domain sizes.
  \item As another point of comparison,---in at most
        \SI{41}{\second}---\textsc{Crane} scales up to domains of sizes \num{e4}
        and \num{3e5} in permutation- and function-counting problems,
        respectively (whereas \textsc{FastWFOMC} already takes longer with
        domains of sizes\dots)
\end{itemize}

@@

% TODO: for references, maybe fix the spelling of Kuzelka's name

>>>>

\newcommand{\Cranetwo}{\textsc{Crane2}}
\newcommand{\Cranebfs}{\textsc{Crane2-BFS}}
\newcommand{\Cranegreedy}{\textsc{Crane2-Greedy}}

@@

\author{%
Ananth K. Kidambi$^1$\and
Guramrit Singh$^1$\and
Paulius Dilkas$^2$\and
Kuldeep S. Meel$^3$ \\
\affiliations
$^1$Indian Institute of Technology Bombay, Mumbai, India\\
$^2$National University of Singapore, Singapore, Singapore\\
$^3$University of Toronto, Toronto, Canada\\
\emails
\{210051002, 210050061\}@iitb.ac.in,
paulius.dilkas@nus.edu.sg,
meel@cs.toronto.edu
}

@@

\todo[inline,caption={}]{
  Add some papers mentioned in:
  \begin{itemize}
    \item very recent work
    \item my previous paper, including:
    \begin{itemize}
      \item other liftable fragments
      \item some more theory papers, e.g., LICS 2018
    \end{itemize}
  \end{itemize}
}

\paragraph{Papers to cite.}
\begin{itemize}
  \item overviews
        \begin{itemize}
          \item lifted probabilistic inference~\cite{DBLP:conf/ecai/Kersting12}
          \item recent overview paper~\cite{DBLP:conf/ijcai/Kuzelka23}
        \end{itemize}
  \item Alternative definition~\cite{DBLP:journals/cacm/GogateD16}
  \item relevant theoretical work~\cite{DBLP:conf/aaai/MalhotraS22}
  \item original domain recursion~\cite{DBLP:conf/nips/Broeck11}
  \item algorithms
        \begin{itemize}
          \item \textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
          \item \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23}
          \item \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21}
          \item \textsc{L2C}~\cite{DBLP:conf/kr/KazemiP16} (similarly to us
                compiles to C++ code, but (probably) doesn't work on as many
                formulas)
          \item approximate~\cite{DBLP:conf/ijcai/BremenK20}
          \item for Markov logic networks~\cite{DBLP:journals/ml/RichardsonD06}
                \begin{itemize}
                  \item \textsc{Magician}~\cite{DBLP:conf/aaai/VenugopalSG15}
                  \item \textsc{Tuffy}~\cite{DBLP:journals/pvldb/NiuRDS11}
                  \item \textsc{Alchemy}~\cite{DBLP:journals/cacm/GogateD16}
                        (same as the alternative definition)
                \end{itemize}
        \end{itemize}
  \item complexity
        \begin{itemize}
          \item liftability~\cite{DBLP:conf/starai/JaegerB12}
          \item hardness for three variables~\cite{DBLP:conf/pods/BeameBGS15}
          \item liftable fragments
                \begin{itemize}
                  \item \Ctwo~\cite{DBLP:journals/jair/Kuzelka21}
                  \item tree axioms~\cite{DBLP:journals/ai/BremenK23}
                  \item linear order axioms~\cite{DBLP:conf/aaai/TothK23}
                  \item some liftable fragments~\cite{DBLP:conf/nips/KazemiKBP16}
                \end{itemize}
        \end{itemize}
  \item applications
        \begin{itemize}
          \item extensions to
                sampling~\cite{DBLP:conf/aaai/WangB0K22,DBLP:conf/lics/WangP0K23}
          \item discovery of combinatorial
                sequences~\cite{DBLP:conf/ijcai/SvatosJT0K23}
          \item conjecturing recurrence
                relations~\cite{DBLP:conf/ilp/BarvinekB0ZK21}
          \item probabilistic logic
                programming~\cite{DBLP:journals/ijar/RiguzziBZCL17} (WFOMC was
                shown to be supreme)
          \item probabilistic databases~\cite{DBLP:journals/debu/GribkoffSB14}
        \end{itemize}
  \item lifted inference in other domains
        \begin{itemize}
          \item constraint satisfaction~\cite{DBLP:journals/jair/TotisDRK23}
          \item answer set programming~\cite{DBLP:journals/ijar/AzzoliniR23}
        \end{itemize}
\end{itemize}

\todo[inline,caption={}]{
  Where to put this?
  \paragraph{Changes to the compilation process.}
  \begin{itemize}
    \item We make the FCGs satisfy the following property. For every domain
          $\Delta$, for every directed path $P$ without Ref nodes, the number of
          atom counting or (generalised) domain recursion or independent partial
          grounding operations on $\Delta$ on $P$ is at most one.
    \item When checking whether recursion is possible between a source and a
          target formula, the target formula (maybe the source formula as well?)
          must not include a domain as well as its subdomain. This might be
          equivalent to the formula being `unshattered'. Such formulas could be
          seen as transitory, and allowing such recursive calls leads to FCGs
          that I cannot make sense of.
  \end{itemize}
}

@@

\paragraph{Three types of logics}
\begin{itemize}
  \item See \cref{tbl:logics} for a detailed comparison. The notation introduced
        in the table is standard for \Ctwo{}, new for \UFO{}, and redefined to
        be more specific for \FO{}.
  \item All three logics are function-free.
  \item Domains are always assumed to be finite.
  \item
        \begin{itemize}
          \item \FO{} is used as the input format for
                \textsc{ForcLift}\footnote{\url{https://github.com/UCLA-StarAI/Forclift}}~\cite{DBLP:conf/ijcai/BroeckTMDR11}
                and its extensions
                \textsc{Crane}\footnote{\url{https://doi.org/10.5281/zenodo.8004077}}~\cite{DBLP:conf/kr/DilkasB23}
                and \Cranetwo{}.
          \item \Ctwo{} is discussed in the literature on
                \textsc{FastWFOMC}~\cite{DBLP:conf/uai/BremenK21} and related
                methods~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/aaai/MalhotraS22}
          \item \UFO{} is the input format supported by a version of
                \textsc{FastWFOMC} obtained directly from the authors. Note that
                the publicly available
                version\footnote{\url{https://comp.nus.edu.sg/~tvanbr/software/fastwfomc.tar.gz}}
                does not support any cardinality constraints.
        \end{itemize}
  \item Note that, in the case of \textsc{ForcLift} and its extensions, support
        for a formula as valid input does not imply that the algorithm will be
        able to compile the formula into a circuit or graph suitable for lifted
        model counting. However, it is known that \textsc{ForcLift} compilation
        is guaranteed to succeed on any \FO{} formula without constants and with
        at most two variables~\cite{DBLP:conf/nips/Broeck11}.
\end{itemize}

@@

\begin{algorithm}[t]
  \caption{The main part of \Cranetwo{} that compiles a formula into a set of
    equations, including the base cases of recursive functions}\label{alg:main}
  \KwIn{formula $\phi$}
  \KwOut{set $\mathcal{E}$ of equations}
  $(\mathcal{E}, \mathcal{F}, \mathcal{D}) \gets \Crane{$\phi$}$\;
  \ForEach{base case $f(\mathbf{x}) \in \FindBaseCases{$\mathcal{E}$}$}{
    $\psi \gets \mathcal{F}(f)$\;
    \ForEach{$i$ such that $x_{i}$ is a constant}{
      $\psi \gets \Propagate{$\psi$, $\mathcal{D}(f, i)$, $x_i$}$\;
    }
    $(\mathcal{E}', \text{\textunderscore}, \text{\textunderscore}) \gets \Crane{$\psi$}$\;
    $\mathcal{E} \gets \mathcal{E} \cup \mathcal{E}'$\;
  }
\end{algorithm}

We write \Crane for the first-order knowledge compilation algorithm of the
original \textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23} that compiles a formula
$\phi$ into:
\begin{itemize}
  \item a set $\mathcal{E}$ of equations,
  \item a map $\mathcal{F}$ from function names to formulas, and
  \item a map $\mathcal{D}$ from function names and parameter indices to
        domains.
\end{itemize}

\todo[inline,caption={}]{
  \begin{itemize}
    \item extend it to work with base cases that are themselves recursive
    \item \textsc{Crane}, \Cranetwo{}, or \Crane (here and elsewhere)?
  \end{itemize}
}

See \cref{alg:main}. The evaluation of base cases is done by simplifying the
clauses and then using \textsc{Crane} to find the base cases. A particular
domain is selected, and the clauses are simplified. Then, \textsc{Crane} is
called on those clauses to evaluate the base cases. After that, we change the
function names and variable to make it consistent with the previous domain to
variable mapping, and append these base cases to the set of equations.

@@

Comparing \Cranetwo{} and \textsc{FastWFOMC} on a larger set of benchmarks is
challenging because there is no automated way to translate a formula in \FO{} or
\Ctwo{} into \UFO{} (or even check if such an encoding is possible).

@@

\paragraph{Setup.}
\begin{itemize}
  \item The experiments were run on an AMD~Ryzen~7~5800H processor with
        \SI{16}{\gibi\byte} of memory and Arch Linux~6.8.2-arch2-1 operating
        system. \textsc{FastWFOMC} was run using Python~3.8.19 with
        Python-FLINT~0.5.0.
  \item The knowledge compilation part of both \textsc{Crane} and \Cranetwo{}
        can be executed using either greedy (similar to \textsc{ForcLift}) or
        breadth-first search. We use both in our experiments, denoting them as
        \Cranegreedy{} and \Cranebfs{}, respectively.
\end{itemize}

@@

\paragraph{Results.}
\begin{itemize}
  \item As shown in \cref{fig:plot}, the runtimes of all compilation-based
        algorithms remain practically constant in contrast to the rapidly
        increasing runtimes of \textsc{FastWFOMC}.
  \item Note that \Cranebfs{} is able to handle more instances than
        \textsc{ForcLift} (e.g., the permutation-counting problem in our
        experiments and other problems in my previous work).
  \item Although the search/compilation part is slower in \Cranetwo{} than in
        \textsc{ForcLift}, the difference is negligible.
  \item The runtimes of three out of four WFOMC algorithms appear constant
        because---for these counting problems and domain sizes---compilation
        time dominates inference time (recall that compilation time is
        independent of domain sizes). Indeed, the maximum inference time of both
        \Cranebfs{} and \Cranegreedy{} across these experiments is
        only \SI{4}{\milli\second}.
  \item The runtimes of \Cranetwo{} have lower variation than those of
        \textsc{ForcLift} because with \textsc{ForcLift} we compile the formula
        anew for each domain size whereas with \Cranetwo{} we compile it once
        and reuse the resulting C++ program for all domain sizes.
  \item As another point of comparison,---in at most
        \SI{41}{\second}---\Cranetwo{} scales up to domains of sizes \num{e4}
        and \num{3e5} in permutation- and function-counting problems,
        respectively (whereas \textsc{FastWFOMC} already takes longer with
        domains of sizes\dots)
\end{itemize}

@@
